### 2015-10-27 docs/ms_communitySynchrony.Rmd
rms bad IBM runs; updates ms filename and caching

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{todonotes}
   - \usepackage[document]{ragged2e}
bibliography: ~/Dropbox/Bibliography/CommunitySynchrony.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
fontsize: 12pt
geometry: margin=1in

---
# Environmental responses synchronize population \newline dynamics in five semi-arid grasslands, but not completely

@@

All organisms exist in variable environments, and those that coexist on similar resources should be expected to respond to the environment in similar ways.
Based on such a proposition, @Loreau2008 suggest that the true null hypothesis is that coexisting species should fluctuate synchronously, not independently, in the absence of temporal niche differenatiation.
Then, the judgment of whether species fluctate synchronously or not depends on the degree to which observed dynamics deviate from perfect synchrony.
In such a light, Houlahan et al.'s (2007) results can be viewed very differently.
However, Houlahan et al.'s (2007) meausure of compensatory dynamics was summed species covariances through time.
Summed covariances have no upper or lower bound, making it difficult to interpret their results based on deviations from perfect synchrony.
In the end, the fact that few communities exhibited negative summed covariance yields little insight as to whether communities tend to be sychronized by environmental forces or not.

Aside from issues relating to the null expectation of species synchrony, disentangling the factors that drive species syncrony is difficult.

What is expected by theory?[^devel]
---------------------------

Loreau and de Mazancourt (2008) derived a statistic to measure community-wide synchrony of per capita population growth rates:

@@

@@

site <- "Idaho"
spp_list <- sort(c("PSSP","HECO","POSE","ARTR"))
num_spp <- length(spp_list)
id_data <- data.frame(quad=NA, year=NA, totCover=NA, species=NA)
for(dospp in 1:num_spp){ #loop through species to read in data
  spp_now <- spp_list[dospp]
  quad_file <- paste("../data/", site,"/",spp_now,"/quadratCover.csv",sep="")
  spp_data <- read.csv(quad_file)
  spp_data$species <- spp_now
  id_data <- rbind(id_data, spp_data)
} #end species looping for raw data
id_data <- id_data[2:nrow(id_data),] #remove first NA row
id_data <- subset(id_data, species!="ARTR") #take out the shrub

@@

```{r time_series_fig, echo=FALSE, include=TRUE, fig.height=7, fig.cap="Observed time series of mean percent cover (averaged over quadrats for each year) and per capita growth rates for each species at each site."}
# out_ks$percent_cover$site <- "Kansas"
# out_id$percent_cover$site <- "Idaho"
# out_mt$percent_cover$site <- "Montana"
# out_nm$percent_cover$site <- "New Mexico"
# out_az$percent_cover$site <- "Arizona"
# cover_data <- rbind(out_ks$percent_cover, out_id$percent_cover,
#                     out_mt$percent_cover, out_nm$percent_cover, 
#                     out_az$percent_cover)
# 
# out_ks$growth_rates$site <- "Kansas"
# out_id$growth_rates$site <- "Idaho"
# out_mt$growth_rates$site <- "Montana"
# out_nm$growth_rates$site <- "New Mexico"
# out_az$growth_rates$site <- "Arizona"
# pgr_data <- rbind(out_ks$growth_rates, out_id$growth_rates,
#                   out_mt$growth_rates, out_nm$growth_rates, 
#                   out_az$growth_rates)
# all_spp <- unique(pgr_data$species)
# color_match <- data.frame(species = all_spp,
#                           color = brewer.pal(11,"Paired"))
# pgr_data <- merge(pgr_data, color_match, by = "species")
# cover_data <- merge(cover_data, color_match, by="species")
# 
# cover_plot <- ggplot(data=cover_data, aes(x=(year+1900), y=tot_cover, color=color))+
#                 geom_line()+
#                 geom_point()+
#                 facet_grid(site~., scales = "free_y")+
#                 guides(color=FALSE)+
#                 ylab("Mean Percent Cover")+
#                 xlab("Year")+
#                 scale_color_manual(values = brewer.pal(11,"Paired"))
# 
# pgr_plot <- ggplot(data=pgr_data, aes(x=(year+1900), y=pgr, color=color))+
#               geom_line()+
#               geom_point()+
#               facet_grid(site~., scales = "free_y")+
#               guides(color=FALSE)+
#               ylab("Per Capita Growth Rate")+
#               xlab("Year")+
#               scale_color_manual(values = brewer.pal(11,"Paired"))
# full_plot <- grid.arrange(cover_plot, pgr_plot, ncol=2)
```

Figures
-------
```{r concept_fig, echo=FALSE, fig.cap="Conceptual figure outlining the effects of interspecific interactions and environmental stochasticity in large populations where demographic stochasticity is effectively absent (A) and in populations of varying size where demographic stochasticity is present (B).", eval=T}
library(png)
library(grid)
img <- readPNG("./prelim_analysis_files/synchrony_conceptual_fig.png")
grid.raster(img)
```

```{r read_data, echo=FALSE, eval=F}
output_list <- readRDS("../results/ipm_comp_nocomp_sims.RDS")
mlist <- melt(output_list)
```

```{r figure_1, echo=FALSE, fig.cap="Observed (vertical dashed lines) and IPM simulated (solid density curves) synchrony of species per capita growth rates based on changes in percent cover.", fig.width=6, fig.height=3, eval=F}
nyrs <- c(max(out_az$stability$num_observations),
          max(out_id$stability$num_observations),
          max(out_ks$stability$num_observations),
          max(out_mt$stability$num_observations),
          max(out_nm$stability$num_observations))

@@

for(do_site in 1:length(sites)){
  full_ipm_outs <- subset(mlist, L2=="ENVINTER")
  site_ipm <- subset(full_ipm_outs, L1==sites[do_site])
  if(sites[do_site]=="Idaho") site_ipm <- subset(site_ipm, Var2!="ARTR")
  species_list <- unique(site_ipm$Var2)
  for(i in 1:boots){
    begin_year <- sample(x = 1:(max(site_ipm$Var1)-nyrs[do_site]), 1)
    end_year <- begin_year+nyrs[1]
    tmp <- subset(site_ipm, Var1 %in% begin_year:end_year)
    tmp2 <- dcast(tmp, Var1~Var2, value.var = "value")

    # caclulate observed growth rates
    # create lagged data frame to only get observed yearly transitions
    lag_df <- tmp2
    lag_df$lagyear <- lag_df$Var1+1
    colnames(lag_df)[2:(num_spp+1)] <- paste0(colnames(lag_df)[2:(num_spp+1)],"_t0")

    # merge the lag df with observed
    rm_col <- which(colnames(lag_df)=="Var1")
    merged_df <- merge(tmp2, lag_df[,-rm_col], by.x = "Var1", by.y="lagyear")
    transitions <- nrow(merged_df)
    obs_gr <- matrix(nrow=transitions, ncol=num_spp)
    name_ids1 <- which(colnames(merged_df) %in% species_list)
    name_ids2 <- which(colnames(merged_df) %in% paste(species_list, "_t0", sep=""))
    for(ii in 1:transitions){
      obs_gr[ii,] <- as.numeric(log(merged_df[ii,name_ids1]/merged_df[ii,name_ids2]))
    }
    synch[i,do_site] <- as.numeric(community.sync(obs_gr)[1])
  }
}

@@

ggplot()+
  geom_line(data=synch_df, aes(x=value, color=Var2), stat="density", adjust=1)+
  geom_vline(data=obs_pgr_df, aes(xintercept=pgr_synch, color=Var2), linetype=2)+
  facet_wrap("Var2")+
  scale_color_manual(values = c("grey45", "steelblue", "slateblue4", "darkorange", "purple"))+
  guides(color=FALSE)+
  theme_bw()
```

```{r figure_2, echo=FALSE, fig.cap="Results from IPM simulations with difference sources of stochasticity present/absent at each site. DEMO = demographic stochasticity only; ENV = environmental stochasticity only; ENVDEMO = both demograhpic and environmental stochasticity.", fig.width=6, eval=F}
colnames(mlist)[1:3] <- c("year", "species", "cover")
sites <- unique(mlist$L1)
sims <- unique(mlist$L2)
pgr_list <- list()
for(dosite in sites){
  tmp_data <- subset(mlist, L1==dosite)
  if(dosite=="Idaho"){
    tmp_data <- subset(tmp_data, species!="ARTR")
  }
  within_site_list <- list()
  for(dosim in sims){
    tmpsim <- subset(tmp_data, L2==dosim)[c("year", "species", "cover")]
    tmpsynch <- get_ipm_synchrony(tmpsim)
    tmp_pgr_synch <- as.numeric(tmpsynch$pgr_synchrony["obs"])
    within_site_list[[dosim]] <- tmp_pgr_synch
  }
  pgr_list[[dosite]] <- within_site_list
}
pgr_synch <- melt(pgr_list)
expected_synch <- table_data[,c(1,3)]
expected_synch$L2 <- "E(SYNCH)"
colnames(expected_synch) <- c("L1", "value", "L2")
expected_synch <- expected_synch[,c("value", "L2", "L1")]
expected_synch[which(expected_synch[,"L1"]=="New Mexico"),"L1"] <- "NewMexico"

@@

print(pgrsynch_plot)

####
####  Compare observed and IPM simulated synchrony
####
nsites <- length(sites)

```

\newpage

Additional Figures
------------------
```{r time_series, eval=FALSE, echo=FALSE, fig.cap="Example time series from the simulations.", eval=F}
ggplot(mlist, aes(x=year, y=(cover*100), color=species))+
  geom_line()+
  facet_grid(L1~L2, scale="free")
```

>>>>

---
layout: 12pt
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage[document]{ragged2e}
bibliography: ~/Dropbox/Bibliography/CommunitySynchrony.bib
csl: components/ecology.csl

## rmarkdown render options
output:
  pdf_document:
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: margin=1in

---
```{r caching, include=FALSE}
library("methods")
library("knitr")
basename <- "ms_communitySynchrony"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)

# PDF-based figures
opts_chunk$set(dev='pdf')
```

# Environmental responses synchronize population dynamics in five semi-arid grasslands, but not completely

@@

All organisms exist in variable environments, and those that coexist on similar resources should be expected to respond to the environment in similar ways.
Based on such a proposition, @Loreau2008a suggest that the true null hypothesis is that coexisting species should fluctuate synchronously, not independently, in the absence of temporal niche differenatiation.
Then, the judgment of whether species fluctate synchronously or not depends on the degree to which observed dynamics deviate from perfect synchrony.
In such a light, Houlahan et al.'s (2007) results can be viewed very differently.
However, Houlahan et al.'s (2007) meausure of compensatory dynamics was summed species covariances through time.
Summed covariances have no upper or lower bound, making it difficult to interpret their results based on deviations from perfect synchrony.
In the end, the fact that few communities exhibited negative summed covariance yields little insight as to whether communities tend to be sychronized by environmental forces or not.

Aside from issues relating to the null expectation of species synchrony, disentangling the factors that drive species syncrony is difficult.
Observational timeseries of species' abundances allows easy quantification of summed covariances [@Houlahan2007] or species synchrony [@Gross2014], but quantifying the relative influence of density-dependence, demographic stochasticity, and environmental stochasticity remains a challenge.
Indeed, to our knowledge, only one other study has attempted to separate the influences of density-dependence and environmental stochasticity [@Mutshinda2009].
@Mutshinda2009 used a statistical technique to decompose synchrony of species' abundances calculate from five time series.
They found that the effect of environmental forcing on temporal dynamics is much stronger than that of interspecific interactions, across all taxa they analyzed [@Mutshinda2009].
However, to link environmental forcing to species synchrony, @Mutshinda2009 examined the correlations among species' environmental responses.

In this paper, we seek to overcome the limitations of previous studies by explicitly testing theory on all the drivers of species synchrony using empirically-based multi-species population models that include the effects of density-dependence, environmental stochasticity, and demographic stochasticity.
Our theoretical framework is that of @Loreau2008a, which defines expectations for the effects of density-dependence, environmental stochasticity, and demographic stochasticity on species synchrony.
We confront theoretical expectations with model-based analyses of community time series from five semi-arid grasslands [@Chu2015].
Our objectives were to (1) reproduce observed species synchrony using multi-species population models, (2) determine the relative influences of environmental stochasticity and interspecific interactions on species synchrony in large populations immune to demographic stochasticity, and (3) determine how dempographic stochasticity augments the influence of the environment and interspecific interactions across a range of populations sizes.
To achieve our objectives, we analyzed spatially-implicit integral projection models (IPMs) and spatially-explicit individually-based models (IBMs) fit to demographic data from each our study sites.
We first show that the models are able to reproduce observed synchrony of species' per capita growth rates and then perturb the models to include/exclude the factors to drive synchrony, as identified by theory.
We used the IPM to model large populations that are immune to demographic stochasticity to isolate the influence of environmental stochasticity; we used the IBM to model populations where demographic stochasticity is present.

Theoretical expectations
------------------------
@Loreau2008a derived a statistic to measure community-wide synchrony of per capita population growth rates:

@@

\normalsize

\pagebreak{}

@@

site <- "Idaho"
spp_list <- sort(c("PSSP","HECO","POSE","ARTR"))
num_spp <- length(spp_list)
id_data <- data.frame(quad=NA, year=NA, totCover=NA, species=NA)
for(dospp in 1:num_spp){ #loop through species to read in data
  spp_now <- spp_list[dospp]
  quad_file <- paste("../data/", site,"/",spp_now,"/quadratCover.csv",sep="")
  spp_data <- read.csv(quad_file)
  spp_data$species <- spp_now
  id_data <- rbind(id_data, spp_data)
} #end species looping for raw data
id_data <- id_data[2:nrow(id_data),] #remove first NA row
# id_data <- subset(id_data, species!="ARTR") #take out the shrub

@@

```{r time_series_fig, echo=FALSE, include=TRUE, fig.height=7, fig.cap="Observed time series of mean percent cover (averaged over quadrats for each year) and per capita growth rates for each species at each site.", eval=FALSE}
# out_ks$percent_cover$site <- "Kansas"
# out_id$percent_cover$site <- "Idaho"
# out_mt$percent_cover$site <- "Montana"
# out_nm$percent_cover$site <- "New Mexico"
# out_az$percent_cover$site <- "Arizona"
# cover_data <- rbind(out_ks$percent_cover, out_id$percent_cover,
#                     out_mt$percent_cover, out_nm$percent_cover, 
#                     out_az$percent_cover)
# 
# out_ks$growth_rates$site <- "Kansas"
# out_id$growth_rates$site <- "Idaho"
# out_mt$growth_rates$site <- "Montana"
# out_nm$growth_rates$site <- "New Mexico"
# out_az$growth_rates$site <- "Arizona"
# pgr_data <- rbind(out_ks$growth_rates, out_id$growth_rates,
#                   out_mt$growth_rates, out_nm$growth_rates, 
#                   out_az$growth_rates)
# all_spp <- unique(pgr_data$species)
# color_match <- data.frame(species = all_spp,
#                           color = brewer.pal(11,"Paired"))
# pgr_data <- merge(pgr_data, color_match, by = "species")
# cover_data <- merge(cover_data, color_match, by="species")
# 
# cover_plot <- ggplot(data=cover_data, aes(x=(year+1900), y=tot_cover, color=color))+
#                 geom_line()+
#                 geom_point()+
#                 facet_grid(site~., scales = "free_y")+
#                 guides(color=FALSE)+
#                 ylab("Mean Percent Cover")+
#                 xlab("Year")+
#                 scale_color_manual(values = brewer.pal(11,"Paired"))
# 
# pgr_plot <- ggplot(data=pgr_data, aes(x=(year+1900), y=pgr, color=color))+
#               geom_line()+
#               geom_point()+
#               facet_grid(site~., scales = "free_y")+
#               guides(color=FALSE)+
#               ylab("Per Capita Growth Rate")+
#               xlab("Year")+
#               scale_color_manual(values = brewer.pal(11,"Paired"))
# full_plot <- grid.arrange(cover_plot, pgr_plot, ncol=2)
```

\pagebreak{}

Figures
-------
```{r concept_fig, echo=FALSE, fig.cap="Conceptual figure outlining the effects of interspecific interactions and environmental stochasticity in large populations where demographic stochasticity is effectively absent (A) and in populations of varying size where demographic stochasticity is present (B).", eval=T}
library(png)
library(grid)
img <- readPNG("./components/synchrony_conceptual_fig.png")
grid.raster(img)
```

```{r read_data, echo=FALSE, eval=T}
output_list <- readRDS("../results/ipm_comp_nocomp_sims.RDS")
mlist <- melt(output_list)
```

```{r figure_1, echo=FALSE, fig.cap="Observed (vertical dashed lines) and IPM simulated (solid density curves) synchrony of species per capita growth rates based on changes in percent cover.", fig.width=6, fig.height=6, fig.width=3, eval=T}
nyrs <- c(max(out_az$stability$num_observations),
          max(out_id$stability$num_observations),
          max(out_ks$stability$num_observations),
          max(out_mt$stability$num_observations),
          max(out_nm$stability$num_observations))

@@

for(do_site in 1:length(sites)){
  full_ipm_outs <- subset(mlist, L2=="ENVINTER")
  site_ipm <- subset(full_ipm_outs, L1==sites[do_site])
  colnames(site_ipm) <- c("year", "species", "cover", "L1", "L2")
  # if(sites[do_site]=="Idaho") site_ipm <- subset(site_ipm, Var2!="ARTR")
  species_list <- unique(site_ipm$species)
  num_spp <- length(species_list)
  for(i in 1:boots){
    begin_year <- sample(x = 1:(max(site_ipm$year)-nyrs[do_site]), 1)
    end_year <- begin_year+nyrs[do_site]
    tmp <- subset(site_ipm, year %in% begin_year:end_year)
    tmp2 <- dcast(tmp, year~species, value.var = "cover")

    # caclulate observed growth rates
    # create lagged data frame to only get observed yearly transitions
    lag_df <- tmp2
    lag_df$lagyear <- lag_df$year+1
    colnames(lag_df)[2:(num_spp+1)] <- paste0(colnames(lag_df)[2:(num_spp+1)],"_t0")

    # merge the lag df with observed
    rm_col <- which(colnames(lag_df)=="year")
    merged_df <- merge(tmp2, lag_df[,-rm_col], by.x = "year", by.y="lagyear")
    transitions <- nrow(merged_df)
    obs_gr <- matrix(nrow=transitions, ncol=num_spp)
    name_ids1 <- which(colnames(merged_df) %in% species_list)
    name_ids2 <- which(colnames(merged_df) %in% paste(species_list, "_t0", sep=""))
    for(ii in 1:transitions){
      obs_gr[ii,] <- as.numeric(log(merged_df[ii,name_ids1]/merged_df[ii,name_ids2]))
    }
    synch[i,do_site] <- as.numeric(community.sync(obs_gr)[1])
  }
}

@@

ggplot()+
  geom_line(data=synch_df, aes(x=value, color=Var2), stat="density", adjust=2)+
  geom_vline(data=obs_pgr_df, aes(xintercept=pgr_synch, color=Var2), linetype=2)+
  facet_wrap("Var2", ncol=1)+
  scale_color_manual(values = c("grey45", "steelblue", "slateblue4", "darkorange", "purple"))+
  guides(color=FALSE)+
  theme_bw()
```

```{r figure_2, echo=FALSE, fig.cap="Results from IPM simulations with difference sources of stochasticity present/absent at each site. DEMO = demographic stochasticity only; ENV = environmental stochasticity only; ENVDEMO = both demograhpic and environmental stochasticity.", fig.width=6, eval=T}
colnames(mlist)[1:3] <- c("year", "species", "cover")
sites <- unique(mlist$L1)
sims <- unique(mlist$L2)
pgr_list <- list()
for(dosite in sites){
  tmp_data <- subset(mlist, L1==dosite)
#   if(dosite=="Idaho"){
#     tmp_data <- subset(tmp_data, species!="ARTR")
#   }
  within_site_list <- list()
  for(dosim in sims){
    tmpsim <- subset(tmp_data, L2==dosim)[c("year", "species", "cover")]
    tmpsynch <- get_ipm_synchrony(tmpsim)
    tmp_pgr_synch <- as.numeric(tmpsynch$pgr_synchrony["obs"])
    within_site_list[[dosim]] <- tmp_pgr_synch
  }
  pgr_list[[dosite]] <- within_site_list
}
pgr_synch <- melt(pgr_list)
expected_synch <- table_data[,c(1,3)]
expected_synch$L2 <- "E(SYNCH)"
colnames(expected_synch) <- c("L1", "value", "L2")
expected_synch <- expected_synch[,c("value", "L2", "L1")]
expected_synch[which(expected_synch[,"L1"]=="New Mexico"),"L1"] <- "NewMexico"

@@

print(pgrsynch_plot)
```

```{r ibm_sims_figure, echo=FALSE, fig.cap="Species synchrony from spatially-explicit IBM simulations with (A,C) and without (B,D) environmental fluctuations and with (A,B) and without (C,D) interspecific interactions. Points represent replicate IBM simulations (out of 20 possible) in which no species went extinct. Lines show linear fits between plot size (proxy for population size) and species synchrony. Demographic stochasticity plays a large role in small populations (small plot sizes) and a smaller role in large populations (large plot sizes).", fig.height=5}
library(png)
library(grid)
img <- readPNG("./components/ibm_sims_fig.png")
grid.raster(img)
```

```{r time_series, eval=FALSE, echo=FALSE, fig.cap="Example time series from the simulations.", eval=F}
ggplot(mlist, aes(x=year, y=(cover*100), color=species))+
  geom_line()+
  facet_grid(L1~L2, scale="free")
```

\pagebreak{}

References
----------
